Analysis-ready VCF at Biobank-scale using Zarr

### 2024-05-03 paper.tex

\subsection{Storing genetic variation data}
Although VCF is the standard format for exchanging genetic variation
data, its limitations both in terms of compression 
and query/compute performance are well 
known~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},
and many methods 
have been suggested to improve on these properties.
Most approaches balance compression with
performance on particular types of queries, 
typically using a command line interface
and outputting VCF text~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
tatwawadi2016gtrac, % GTRAC
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
lefaive2021sparse, % SAVVY
wertenbroek2022xsi,% XSI
zhang2023gbc}. %GBC
Several specialised algorithms for compressing 
the genotype matrix (i.e., just the genotype calls without additional
VCF information) have been proposed
\citep{deorowicz2013genome, %TGC
deorowicz2019gtshark, %GTShark
deorowicz2021vcfshark, % VCFShark
dehaas2024genotype} %  GRG
most notably the Positional
Burrows--Wheeler Transform (PBWT)~\citep{durbin2014efficient}.
See~\citep{mcvean2019linkage} for a review of the techniques
employed in genetic data compression.

VCF is a text-based format in which each line
describes observations
and metadata for a single variant (usually a position on the genome
where variation occurs in the samples)~\citep{danecek2011variant}. 
VCF is therefore a \emph{row-wise} format, i.e., where all of the 
data for a record is stored continguously. BCF is similarly 
row-wise, as are the majority of proposed alternative storage formats.
Row-wise storage makes retrieving all information
for a given record straightforward and efficient, and 
allows data to be easily streamed through Unix pipelines (see Discussion).

Row-wise storage works well when records are relatively small, 
or when we typically want to analyse each record in its entirity.
When we want to analyse only a subset of a record,
row-wise storage can be inefficient because we will usually need to
retrieve more information than required from storage. In the case 
of VCF (and BCF) where records are not of a fixed size and 
are almost always compressed in blocks, accessing any information
for all rows means retrieving and decompressing \emph{all} 
information from all rows.
The alternative to row-wise storage is \emph{columnar} storage:
instead of grouping together all the fields for a record,
we group together all the records for a given field.
Columnar storage has become standard [very quick summary of columnar
storage with some refs].
Although columnar methods have been successfully used in alignment 
% TODO check
storage~\citep{bonfield2014scramble,bonfield2022cram},
the only existing method using columnar 
storage for genetic variation data is 
SeqArray~\citep{zheng2017seqarray,zheng2012high} to our knowledge.

@@

>>>>

\subsection{Storing genetic variation data}
Although VCF is the standard format for exchanging genetic variation
data, its limitations both in terms of compression 
and query/compute performance are well 
known~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},
and many methods 
have been suggested to improve on these properties.
Most approaches balance compression with
performance on particular types of queries, 
typically using a command line interface
and outputting VCF text~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
tatwawadi2016gtrac, % GTRAC
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
lefaive2021sparse, % SAVVY
wertenbroek2022xsi,% XSI
zhang2023gbc}. %GBC
Several specialised algorithms for compressing 
the genotype matrix (i.e., just the genotype calls without additional
VCF information) have been proposed
\citep{deorowicz2013genome, %TGC
deorowicz2019gtshark, %GTShark
deorowicz2021vcfshark, % VCFShark
dehaas2024genotype} %  GRG
most notably the Positional
Burrows--Wheeler Transform (PBWT)~\citep{durbin2014efficient}.
See~\citep{mcvean2019linkage} for a review of the techniques
employed in genetic data compression.
[FIXME need to work SeqArray~\citep{zheng2017seqarray,zheng2012high}
in here.]

VCF is a text-based format in which each line
describes observations
and metadata for a single variant~\citep{danecek2011variant}. 
VCF is therefore a \emph{row-wise} format, i.e., where all of the 
data for a record is stored continguously. BCF (a packed binary 
representation) is similarly 
row-wise, as are the majority of proposed alternative storage formats.
Row-wise storage makes retrieving all information
for a given record straightforward and efficient, and 
allows data to be easily streamed through Unix pipelines (see Discussion).

Row-wise storage works well when records are relatively small, 
or when we typically want to analyse each record in its entirity.
When we want to analyse only a subset of a record,
row-wise storage can be inefficient because we will usually need to
retrieve more information than required from storage. In the case 
of VCF (and BCF) where records are not of a fixed size and 
are almost always compressed in blocks, accessing any information
for a set of rows means retrieving and decompressing \emph{all} 
information from these rows.
The alternative to row-wise storage is \emph{columnar} storage:
instead of grouping together all the fields for a record,
we group together all the records for a given field.
Columnar storage formats such as Parquet~\citep{parquet2024},
makes retrieving particular columns much 
more efficient and can lead to substantially better compression.
While columnar techniques have been successfully applied 
alignment
% TODO check
% Jeff: OK with the ADAM reference in here? You did some parquet based
% alignment storage, right?
storage~\citep[e.g.][]{bonfield2014scramble,nothaft2015rethinking,bonfield2022cram},
use of technologies such as Parquet for 
storing and analysing variation data have had limited
success~\citep{boufea2017managing,fan2020variant}.
Directly mapping VCF data to a columnar layout, in which there is a 
a column for the genotypes (and other per-sample QC metrics) 
for each sample, scales poorly. Firstly, it leads to a large number 
of columns (millions, at Biobank scale), which techologies such 
as Parquet are not designed for. Secondly, compression performance 
within columns is poor because there is much less similarity among
the genotypes for a given sample than there is among the genotypes
for a given variant. [TODO check this - reference?]

@@

[TODO intro passage here on array-based or ``tensor'' storage. Separate
out the different interleaved matrices with the VCF, and store them
in regular chunked grids.]
